---
title: "R Bridge - Week 2 Assignment"
author: "Leland Randles"
date: "July 24, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## lr_combo(n, r):  A Function to Compute "n choose r"

* R provides built-in functions for:
    + computing number of combinations - choose(n, r)
    + computing factorials - factorial(x)
* The formula for "n choose r" is n! / ((n - r)! * r!)
* The assignment is to write a function which accomplishes the same thing as the choose(n, r) function without using the aforementioned built-in functions

### Overview of approach and assumptions

The lr_combo(n, r) function assumes that n and r are both whole numbers. This is a departure from the built-in R functions.  For example, factorial(x) will return results for a decimal value:

```{r}
factorial(2.2)
```

And choose(n, r) will accept a decimal value for n:

```{r}
choose(10.2, 3)
```

Aside from this assumption, lr_combo(n, r) should behave exactly the same as choose(n, r).

### Outline of code:
1. Test inputs:
    + Check to see if n and r are whole numbers, send message if they are not
    + Check to see if n is greater than or equal to r, send message if it is not
2. Create factorial function
3. Compute "n choose r"

```{r}
lr_combo <- function(n, r) {
        ti <- TRUE
        #Make sure n and r are whole numbers
        if (is.numeric(n) == FALSE | n < 0 | floor(n) != n) {
                message("n must be a whole number.")
                ti <- FALSE
        }
        if (is.numeric(r) == FALSE | r < 0 | floor(r) != r) {
                message("r must be a whole number.")
                ti <- FALSE
        }
        #Make sure n is greater than or equal to r
        if (r > n) {message("n must be greater than or equal to r")
                ti <- FALSE}
        
        #Factorial function
        lr_fact <- function(x) {
                if (x <= 1) {return(1)
                } else 
                {y <- 1
                for (i in 2:x) {
                        y <- y * i
                } 
                return(y)
                }
        }
                
        #Compute n choose r using combination formula
        if (ti == TRUE) {
                c <- lr_fact(n) / (lr_fact(n - r) * lr_fact(r))
                print(c)
        }
}
```


### Test Cases


**Test 1:  n is negative**
```{r}
lr_combo(-5, 3)
```
lr_combo correctly informs user that n is not a whole number and r is greater than n


**Test 2:  n is a decimal**
```{r}
lr_combo(5.2, 2)
```
lr_combo correctly informs user that n is not a whole number


**Test 3:  r is negative**
```{r}
lr_combo(5, -3)
```
lr_combo correctly informs user that r is not a whole number


**Test 4:  r is a decimal**
```{r}
lr_combo(5, 2.8)
```
lr_combo correctly informs user that r is not a whole number


**Test 5:  r is greater than n**
```{r}
lr_combo(6, 8)
```
lr_combo correctly informs user that r is greater than n


**Test 6:  compute 5 choose 3 - expected answer is 10**
```{r}
lr_combo(5, 3)
```
lr_combo returns the expected value 10


**Test 7:  compute 14 choose 6 - expected answer is 3003**
```{r}
lr_combo(14, 6)
```
lr_combo returns the expected value 3003


**Test 8:  compute 11 choose 8 - expected answer is 165**
```{r}
lr_combo(11, 8)
```
lr_combo returns the expected value 165


**Test 9:  compute 52 choose 3 - expected answer is 22100**
```{r}
lr_combo(11, 8)
```
lr_combo returns the expected value 22100
